Dark Control:\\ A Unified Account of Default Mode Function\\ by Control Theory and Reinforcement Learning

### 2017-05-03 paper/paper.tex
itemr1

This dark matter of brain physiology \citep{raichle2006brain} begs the question
of the biological purpose underlying DMN activity.
What has early been described as the ``stream of consciousness''
in psychology \citep{james1890principles}
found a potential neurobiological manifestation
in the DMN
\citep{shul1997, raichle2001pnas}.
We propose that this set of some of the most advanced regions in
the human association cortex \citep{mesulam1998sensation, margulies2016situating}
are responsible for higher-order control of human behavior.
Our functional account follows the notion of
``a hierarchy of brain systems with
the DMN at the top and the salience and dorsal attention systems
at intermediate levels, above thalamic and unimodal sensory
cortex'' \citep{carhart2010default}.

\subsection{Towards a formal account of default mode function: higher-order control of the organism}
The human DMN is responsible for a large part of the baseline neural activity,
which typically decreases when engaged in psychological experiments.
The standard mode of
neural information maintenance and manipulation has
been argued to mediate evolutionarily conserved functions
\citep{brown1914nature, binder1999conceptual, buzsaki2006rhythms}.
Today, many psychologists and neuroscientists believe that
the DMN implements some form of
probabilistic estimation of past, hypothetical, and
future events
\citep{fox2005, hassabis2007patients, schacter2007remembering, binder2009,
randy2008, spreng2009common}.
This brain network
might have emerged to continuously predict environmental events using
mental imagery as an evolutionary advantage.
%
However, information processing in the DMN has also repeatedly
been shown to directly impact human behavior. Goal-directed task performance
improved with decreased activity in default mode regions \citep{weiss2006}
and increased DMN activity was linked to more task-independent,
yet sometimes useful thoughts
\citep{mason2007, seli2016mind}.
%
Gaining insight into DMN function is
particularly challenging because
this network appears to
simultaneously impact perception-action cycles and
maintain probabilistic contemplations
across time, space, and content domains
\citep{boyer2008evolutionary}.

The present work adopts the perspective of
a human agent faced with the choice of the next actions
and guided by outcomes
of really happened, hypothetically imagined, and
expected futures to optimize behavioral performance.
Formally, a particularly attractive framework
to describe, quantify, and predict intelligent systems, such as the brain,
is proposed to be the combination of control theory and
reinforcement learning (RL).
An intelligent agent improves its interaction with the environment
by continuously updating its computation of value estimates and action predispositions
through integration of feedback outcomes.
Here,
\textit{control} refers to the influence that an agent exerts when interacting
with the environment to encourage preferred states.

Psychologically,
it is known that, the more the world is easy to predict,
the more human mental activity
becomes detached from the actual sensory environment
\citep{antrobus1966studies, pope1978regulation}.
Conversely,
the more the ongoing executed task is unknown and unpracticed,
the less stimulus-independent thoughts occur
\citep{filler1973daydreaming, teasdale1995stimulus, christoff2016mind}.
Without requiring explicit awareness by the agent,
these ``offline'' processes may contribute
to optimizing control of the organism.
We formalize
a policy matrix to capture the space of possible actions that can be performed
on the environment given the current state. A value function
maps environmental objects and events (i.e., states) to expected rewards.
Switching between states reduces to a sequential processing model.
Informed by outcomes of really performed action,
the DMN dynamics are constantly adapted in feedback loops
shaped by prediction error.
A RL account of DMN function can naturally embed human behavior
into the tension between exploitative action with immediate gains and
exploratory action with longer-term gratification.

We argue that DMN implication in many
advanced cognitive processes in humans
can be recast as prediction error minimization
based on probabilistic mental simulations,
thus maximizing action outcome across multiple time scales.
Such a purposeful optimization objective
may be solved by a stochastic approximation
based on a brain implementation of Markov Chain Monte Carlo (MCMC) sampling
\citep{tenenbaum2011grow}.
Even (necessarily imperfect) memory recall,
random day-time mind-wandering, and
seemingly arbitrary dreams during sleep
may provide meaningful building blocks to iteratively improve
the predictive DMN machinery to optimize the behavior of the organism.
%
It has indeed been proposed that
the human brain's energy budget is largely dedicated to
``the development and maintenance of [a]
probabilistic model of anticipated events''
\citep{raichle2005intrinsic}.
This idea is invigorated by
empirical evidence from
neuroscience experiments \citep{kording2004bayesian, fiser2004small}.
The present paper proposes a
process model that satisfies this contention.

@@

Normal and disturbed metabolic fluctuations in the
human PMC have been closely related to
changes of conscious awareness \citep{cavanna2006precuneus}.
Indeed,
the PMC matures relatively late (i.e., myelination) during postnatal development in monkeys
\citep{goldman1987development}, which is generally considered to
be a sign of evolutionary sophistication \citep{flechsig1920}.
%
This DMN node has long been speculated to
reflect constant computation of
environmental statistics and its internal representation
as an inner ``mind's eye'' \citep{cavanna2006precuneus, leech_pcc2014}.
For instance, B\'alint's syndrome is a neurological disorder of conscious
awareness that results from medial damage in the parietal cortex
\citep{balint1909seelenlahmung}.
Patients are plagued by an
inability to combine various individual features of the visual
environment into an integrated whole (i.e., simultanagnosia)
as well as an inability to direct action towards
currently unattended environmental objects
(i.e., optic ataxia).
This can be viewed as a high-level impairment in gathering
information about alternative objects (i.e., exploration) as well as
leveraging these environmental opportunities (i.e., exploitation).
Congruently,
the human PMC was coupled in two functional connectivity analyses
\citep{bzdok2015subspecialization}
with the amygdala, involved in significance evaluation, and
the nucleus accumbens (NAc), involved in reward evaluation.
Specifically, among all parts of the PMC,
the ventral posterior cingulate cortex was
most connected to the laterobasal
nuclei group of the amygdala
\citep{bzdok2015subspecialization}.
This amygdalar subregion has been proposed to
continuously scan environmental input
for biological significance assessment.

The putative role of the PMC in continuous information integration of
environmental relevance
and ensuing top-level guidance of action on the environment is supported
by many neuroscience experiments.
Electrophysiological recordings in animals implicated PMC neurons in
strategic decision making \citep{pearson2009neurons},
risk assessment \citep{mccoy2005risk},
outcome-dependent behavioral modulation \citep{hayden2009electrophysiological},
as well as approach-avoidance behavior
\citep{vann2009does}.
Neuron spiking activity in the PMC allowed distinguishing
whether a monkey would persue an exploratory or exploitative
behavioral strategy during food foraging \citep{pearson2009neurons}.
Further, single-cell recordings in the monkey PMC
demonstrated this brain region's sensitivity to
subjective target utility \citep{mccoy2005risk} and integration
across individual decision-making instances \citep{pearson2009neurons}.
This DMN node encoded the
preference of or aversion to options with uncertain reward outcomes
and its spiking activity was more associated with
subjectively perceived relevance of a chosen object
than by its actual value,
based on an ``internal currency of value'' \citep{mccoy2005risk}.
In fact, direct stimulation of PMC neurons
promoted exploratory actions,
which would otherwise be shunned \citep{hayden2008stim}.
Graded changes in firing rates of PMC neurons
indicated changing choices in upcoming trials and their neural patterns were
distinct from neuronal spike firings that indicated choosing either option.
Similarly in humans,
the DMN has been shown to gather and integrate information
over different parts of auditory narratives in an fMRI study
\citep{simony2016dynamic}.

@@

Comparing to the dmPFC,
the vmPFC probably subserves
subjective value evaluation and
risk estimation of relevant environmental stimuli.
This DMN node may
orchestrate adaptive behavior by bottom-up-driven
processing of “what matters now”,
probably drawing on sophisticated value representations
\citep{doherty2015structure}.
Quantitative lesion findings across 344 human individuals confirmed
a substantial impairment in value-based action choice
\citep{glascher2012lesion}.
Indeed,
this DMN node is preferentially connected with reward-related and limbic regions.
The vmPFC is well known to have strong axonal connections
with the NAc
in axonal tracing studies in monkeys \citep{haber1995orbital}.
Congruently, the gray-matter volume of the vmPFC and NAc
correlated with indices of value-guided behavior and reward attitudes
in humans
\citep{lebreton2009automatic}.
NAc activity is thought to reflect reward prediction signals
from dopaminergic neurotransmitter pathways
\citep{schultz1998predictive}
that not only channel action towards basic survival needs but also
subserve RL in humans more broadly \citep{doherty2015structure}.
Consistently, diffusion MRI tractography in humans and monkeys
\citep{croxson2005quantitative}
quantified the NAc to
be more connected to the vmPFC than dmPFC in both species.
Two different functional connectivity analyses in humans also strongly connected
the vmPFC with the NAc, hippocampus (HC),
and PMC \citep{bzdok2015subspecialization}.
%
In line with these connectivity findings in animals and humans,
the vmPFC is often proposed to represent triggered
emotional and motivational states \citep{damasio1996somatic}.
Such real or imagined arousal states could be mapped in the vmPFC
as a bioregulatory disposition influencing cognition
and decision making.
In neuroeconomic studies of human decision making,
the vmPFC consistently reflects an individual’s subjective
value estimates
\citep{behrens2008associative}.
This may be why performance within and across participants
was related to state encoding in the vmPFC \citep{Schuck2016}.
Such a ``cognitive map'' of the action space was argued to encode
the current task state even when states are unobservable from the sensory environment.

\subsection{The hippocampus: memory, space, and experience replay}
The DMN midline is probably closely linked
with the HC in the medial temporal lobe \citep{vincet2006, shannon2013morning} \textemdash
a region long known to be involved in
memory operations and spatial navigation in animals and humans.

@@

The TPJ in the right-hemispheric DMN (RTPJ)
has been shown to be closely related to
multi-sensory prediction and prediction error signaling.
It is probably central for
action initiation during goal-dircted psychological tasks and for
sensorimotor control by integrating multi-sensory attention
\citep{corbetta2002control}.
Involvement of this DMN node was repeatedly reported in
multi-step action execution \citep{hartmann2005takes},
visuo-proprioceptive conflict \citep{Balslev2005}, and
detection of environmental changes across
visual, auditory, or tactile stimulation in an fMRI experiment
\citep{downar2000multimodal}.
Direct electrical stimulation of the human
RTPJ during neurosurgery was associated with altered perception
and stimulus awareness \citep{blanke2002neuropsychology}.
%
It was argued that the RTPJ encodes actions and ensuing outcomes,
without necessarily relating those to value estimation
\citep{liljeholm2013neural, hamilton2008action,
jakobs2009effects}.
Additionally, neural activity in the RTPJ has been proposed to reflect
stimulus-driven attentional reallocation to
self-relevant and unexpected sources of information
as a ‘‘circuit breaker’’ that recalibrates control and maintenance brain networks
\citep{bzdok2013tpj, corbettashul2008}.
% (Bzdok et al., 2013; Corbetta et al., 2008).
Indeed, patients with RTPJ damage have particular difficulties
with multi-step actions \citep{hartmann2005takes}.
In the face of large discrepancies between actual and previously predicted
environmental events the RTPJ acts as a potential switch between
externally-oriented mind sets focussed on the
sensory environment and internally-oriented mind sets focussed
on self-relevant mental scene construction.
For instance, temporally induced RTPJ damage in humans diminished the
impact of predicted intentions of other individuals
\citep{young2010disruption},
a capacity believed to be enabled by the DMN.
The RTPJ might hence be an important relay that shifts away
from the ‘‘internally directed’’ baseline processes
to, instead, deal with (unexpected) environmental objects and events.
% (TODO: slow endogeneous changes in state transitions by the policy and value matrices)

The TPJ in the left-hemispheric DMN (LTPJ),
in turn, has a close relationship to Wernicke's area
involved in semantic processes, such as in spoken and written language.
Neurological patients with damage in Wernicke's area
have a major impairment of language comprehension
when listening to others or reading a book.
Patient speech
preserves natural rhythm and normal syntax, yet the
voiced sentences lack meaning (i.e., aphasia).
Abstracting from speech interpretations in linguistics
and neuropsychology,
the LTPJ probably mediates access to and integration of world knowlege,
such as required during action considerations
\citep{binder2011neurobiology, seghier2013angular}.
For instance, LTPJ damage in humans also entails problems in recognizing
others' pantomimed action towards objects
without obvious relation to processing explicit language content
\citep{varney1987locus}.
% (Varney and Damasio 1987; Rothi et al. 1991; Buxbaum et al. 2005).
%
Inner speech also hinges on knowledge recall
about the physical and social world.
Indeed,
the internal production of
verbalized thought (``language of the mind'') was closely related to the LTPJ
in a pattern analysis of brain volume
\citep{geva2011neural}.
Further,
episodic memory recall and imagination strongly draw on
reassembling world knowledge.
Isolated building blocks of world structure probably get rebuilt
in internally constructed mental scenarios that
guide present action choice,
weigh hypothetical possibilities, and forecast the future.
%
The LTPJ may hence facilitate the automated prediction of events
by incorporating experience-derived building blocks of world regularities
into ongoing action, planning, and problem solving.
% world -> policy matrix driven computation
% self -> value matrix driven computation

@@

\begin{wrapfigure}{r}{.4\linewidth}
  \caption{\textbf{Reinforcement learning in a nutshell.} Given the current state of the environment,
    the agent takes an action (by following some policy). This yields a reward and an observed
    state transition of the environment. This process goes on until interrupted or a goal state is reached.}
  \centering
  \includegraphics[width=1.\linewidth]{rl.pdf}
  \label{fig:rl}
\end{wrapfigure}

@@

In artificial intelligence and machine learning, a popular computational model for
multi-step decision processes in such an environment are
\textit{Markov decision processes (MDPs)}\citep{sutton1998reinforcement}.
% For the remainder of this paper, for the sake of simplicity
% we use the abbreviation ``MDP'' to refer to the special case
% of \textit{Partially Observable Markov Decision Processes}.
A MDP operationalizes a sequential decision process
in which it is assumed that environment dynamics are determined by a Markov process,
but the agent cannot directly observe the underlying state.
Instead, the agent tries to optimize a \textit{subjective} reward
signal (i.e., is likely to be different for another agent in the same state).
This is performed alongside each transition in the environments state space, and maintains a probability distribution over the set of possible
states, based on a set of observations and observation probabilities. This is a minimal amount of assumptions that can be made about an environment,
and is characteristic of so-called \textit{model-free RL}. Model-free RL has had great success in real-world problems including robotics
\citep{abbeel2004,andrewng}, computer game play, \citep{mnih2015,silver2016mastering}, etc., just to name a few.
We propose to use such a model as formal basis to model DMN function.

@@

Since an action executed in the present moment might have repercussions in the far future, it turns out that the
quantity to optimize is not the instantaneous rewards $r(s, a)$, but a
cumulative reward estimate which takes into account expected reward in the future.
A common approach to
modeling this accumulation is the time-discounted \textit{cumulative reward}% function
\begin{equation}
  \label{eq:cumr}
  G^\pi = \sum_{t=0}^{\infty}\gamma^{t}R_t = R_0 + \gamma R_1 + \gamma^2 R_2 + \ldots + \gamma^tR_t + \ldots
\end{equation}
This random variable\footnote{Random as it depends both on the environment's dynamic and the
  policy $\pi$ being played (which can be stochastic).}  measures the cumulative reward of
following a policy $\pi$.
%
Note that value buffering may be realized in the vmPFC.
This DMN node has direct connections to
to the NAc, known to be involved in reward evaluation,
and
the amygdala, known to be involved in evaluation of biological significance.

The goal of the RL agent is then to update this action policy in order
to maximize $G^\pi$ on average (cf. below). In \eqref{eq:cumr},
the definition of cumulative reward $G^\pi$,
the constant $\gamma$ $(0 \le \gamma < 1$) is the \textit{reward discount factor},
viewed to be specific an agent.
On the one hand,
setting $\gamma = 0$ yields perfectly hedonistic behavior.
An agent with such a shortsighted horizon is exclusively
concerned with immediate rewards.
This is however not compatible with coordinated planning if long-term goal that is
potentially subserved by neural activity in the DMN.
% \footnote{In which case we can take $T \rightarrow +\infty$, provided the reward function $r$ is bounded.}.
% The reward discount factor
On the other hand,
setting $0 < \gamma < 1$ allows a learning process to arise.
A positive $\gamma$ can be seen as calibrating risk-seeking trait of the intelligent agent,
that is, the behavioral predispositions related to trading longer delays for higher rewards.
Such an agent puts relatively more emphasis on rewards expected in
a longer-term future and pays less attention to
rewards expected at a later time. Precisely, rewards that are not expected to come within $\tau := 1/(1 - \gamma)$ time steps from the present point are disregarded.
This reduces the variance of expected rewards accumulated across
considered action cascades by limiting the depth of the search tree.
Given that there is more uncertainty in the farsighted future,
it is important to appreciate that a stochastic policy estimation
is more advantageous in many RL settings.
% For instance, a deterministic strategy in playing rock-paper-scissors can
% be easily exploited and a uniformly random choice is optimal.

@@

% \subsubsection{The Bellman equation:  characterizing optimal behavior}
% Q-learning (Watkins \& Dayan, 1992) generates a
% \textit{greedy policy}
\subsubsection{Optimal behavior and the Bellman temporal consistency}

>>>>

This dark matter of brain physiology \citep{raichle2006brain} begs the question
of the biological purpose underlying DMN activity.
What has early been described as the ``stream of consciousness''
in psychology \citep{james1890principles}
found a potential neurobiological manifestation
in the DMN
\citep{shul1997, raichle2001pnas}.
We propose that this set of some of the most advanced regions in
the association cortex \citep{mesulam1998sensation, margulies2016situating}
are responsible for higher-order control of human behavior.
Our functional account follows the notion of
``a hierarchy of brain systems with
the DMN at the top and the salience and dorsal attention systems
at intermediate levels, above thalamic and unimodal sensory
cortex'' \citep{carhart2010default}.

\subsection{Towards a formal account of default mode function: higher-order control of the organism}
The human DMN is responsible for a large part of the baseline neural activity,
which typically decreases when engaged in psychological experiments
\citep{raichle_baseline}.
The standard mode of
neural information maintenance and manipulation has
been argued to mediate evolutionarily conserved functions
\citep{brown1914nature, binder1999conceptual, buzsaki2006rhythms}.
Today, many psychologists and neuroscientists believe that
the DMN implements some form of
probabilistic estimation of past, hypothetical, and
future events
\citep{fox2005, hassabis2007patients, schacter2007remembering, binder2009,
randy2008, spreng2009common}.
This brain network
might have emerged to continuously predict the environment using
mental imagery as an evolutionary advantage \citep{suddendorf2007evolution}.
%
However, information processing in the DMN has also repeatedly
been shown to directly impact human behavior. Goal-directed task performance
improved with decreased activity in default mode regions \citep{weiss2006}
and increased DMN activity was linked to more task-independent,
yet sometimes useful thoughts
\citep{mason2007, seli2016mind}.
%
Gaining insight into DMN function is
particularly challenging because
this network appears to
simultaneously impact perception-action cycles and
maintain probabilistic contemplations
across time, space, and content domains
\citep{boyer2008evolutionary}.

The present work adopts the perspective of
a human agent faced with the choice of the next actions
and guided by outcomes
of really happened, hypothetically imagined, and
expected futures to optimize behavioral performance.
Formally, a particularly attractive framework
to describe, quantify, and predict intelligent systems, such as the brain,
is proposed to be the combination of control theory and
reinforcement learning (RL).
An intelligent agent improves its interaction with the environment
by continuously updating its computation of value estimates and action predispositions
through integration of feedback outcomes.
Henceforth,
\textit{control} refers to the influence that an agent exerts when interacting
with the environment to reach preferred states.

Psychologically,
the more the ongoing executed task is unknown and unpracticed,
the less stimulus-independent thoughts occur
\citep{filler1973daydreaming, teasdale1995stimulus, christoff2016mind}.
Conversely,
it is known that, the more the world is easy to predict,
the more human mental activity
becomes detached from the actual sensory environment
\citep{antrobus1966studies, pope1978regulation}.
Without requiring explicit awareness by the agent,
these ``offline'' processes may contribute
to optimizing control of the organism.
We formalize
a \textit{policy matrix} to capture the space of possible actions that can be performed
on the environment given the current state. A \textit{value function}
maps environmental objects and events (i.e., states) to expected rewards.
Switching between states reduces to a sequential processing model.
Informed by outcomes of really performed action,
the DMN dynamics could be constantly adapted in feedback loops
shaped by prediction error.
Such an RL account of DMN function can naturally embed human behavior
into the tension between exploitative action with immediate gains and
exploratory action with longer-term gratification.

We argue that DMN implication in many
advanced cognitive processes in humans
can be recast as prediction error minimization
based on probabilistic mental simulations,
thus maximizing action outcome across multiple time scales.
Such a purposeful optimization objective
may be solved by a stochastic approximation
based on a brain implementation of Markov Chain Monte Carlo (MCMC) sampling
\citep{tenenbaum2011grow}.
Even necessarily imperfect memory recall,
random day-time mind-wandering, and
seemingly arbitrary dreams during sleep
may provide blocks of pseudo-experience to iteratively
optimize the behavior of the organism.
%
It has indeed been proposed that
the human brain's energy budget is largely dedicated to
``the development and maintenance of [a]
probabilistic model of anticipated events''
\citep{raichle2005intrinsic}.
This idea is invigorated by
empirical evidence from
neuroscience experiments \citep{kording2004bayesian, fiser2004small}.
The present paper proposes a
process model that satisfies this contention.

@@

Normal and disturbed metabolic fluctuations in the
human PMC have been closely related to
changes of conscious awareness \citep{cavanna2006precuneus}.
Indeed,
the PMC matures relatively late (i.e., myelination) during postnatal development in monkeys
\citep{goldman1987development}, which is generally considered to
be a sign of evolutionary sophistication \citep{flechsig1920}.
%
This DMN node has long been speculated to
reflect constant computation of
environmental statistics and its internal representation
as an inner ``mind's eye'' \citep{cavanna2006precuneus, leech_pcc2014}.
For instance, B\'alint's syndrome is a neurological disorder of conscious
awareness that results from medial damage in the parietal cortex
\citep{balint1909seelenlahmung}.
Patients are plagued by an
inability to combine various individual features of the visual
environment into an integrated whole (i.e., simultanagnosia)
as well as an inability to direct action towards
currently unattended environmental objects
(i.e., optic ataxia).
This can be viewed as a high-level impairment in gathering
information about alternative objects (i.e., exploration) as well as
leveraging these environmental opportunities (i.e., exploitation).
Congruently,
the human PMC was coupled in two functional connectivity analyses
\citep{bzdok2015subspecialization}
with the amygdala, involved in significance evaluation, and
the nucleus accumbens (NAc), involved in reward evaluation.
Specifically, among all parts of the PMC,
the ventral posterior cingulate cortex was
most connected to the laterobasal
nuclei group of the amygdala
\citep{bzdok2015subspecialization}.
This amygdalar subregion has been proposed to
continuously scan environmental input
for biological significance assessment
\citep{trust_attract}.

The putative role of the PMC in continuous information integration of
environmental relevance
and ensuing top-level guidance of action on the environment is supported
by many neuroscience experiments.
Electrophysiological recordings in animals implicated PMC neurons in
strategic decision making \citep{pearson2009neurons},
risk assessment \citep{mccoy2005risk},
outcome-dependent behavioral modulation \citep{hayden2009electrophysiological},
as well as approach-avoidance behavior
\citep{vann2009does}.
Neuron spiking activity in the PMC allowed distinguishing
whether a monkey would persue an exploratory or exploitative
behavioral strategy during food foraging \citep{pearson2009neurons}.
Further, single-cell recordings in the monkey PMC
demonstrated this brain region's sensitivity to
subjective target utility \citep{mccoy2005risk} and integration
across individual decision-making instances \citep{pearson2009neurons}.
This DMN node encoded the
preference for or aversion to options with uncertain reward outcomes
and its spiking activity was more associated with
subjectively perceived relevance of a chosen object
than by its actual value,
based on an ``internal currency of value'' \citep{mccoy2005risk}.
In fact, direct stimulation of PMC neurons
promoted exploratory actions,
which would otherwise be shunned \citep{hayden2008stim}.
Graded changes in firing rates of PMC neurons
indicated changing choices in upcoming trials and their neural patterns were
distinct from neuronal spike firings that indicated choosing either option.
Similarly in humans,
the DMN has been shown to gather and integrate information
over different parts of auditory narratives in an fMRI study
\citep{simony2016dynamic}.

@@

Comparing to the dmPFC,
the vmPFC probably subserves
subjective value evaluation and
risk estimation of relevant environmental stimuli.
This DMN node may
orchestrate adaptive behavior by bottom-up-driven
processing of “what matters now”,
probably drawing on sophisticated value representations
\citep{rolls_OFC, doherty2015structure}.
Quantitative lesion findings across 344 human individuals confirmed
a substantial impairment in value-based action choice
\citep{glascher2012lesion}.
Indeed,
this DMN node is preferentially connected with reward-related and limbic regions.
The vmPFC is well known to have strong axonal connections
with the NAc
in axonal tracing studies in monkeys \citep{haber1995orbital}.
Congruently, the gray-matter volume of the vmPFC and NAc
correlated with indices of value-guided behavior and reward attitudes
in humans
\citep{lebreton2009automatic}.
NAc activity is thought to reflect reward prediction signals
from dopaminergic neurotransmitter pathways
\citep{schultz1998predictive}
that not only channel action towards basic survival needs but also
subserve reward processing, and thus perhaps RL,
in humans more broadly \citep{doherty2015structure}.
Consistently, diffusion MRI tractography in humans and monkeys
\citep{croxson2005quantitative}
quantified the NAc to
be more connected to the vmPFC than dmPFC in both species.
Two different functional connectivity analyses in humans also strongly connected
the vmPFC with the NAc, hippocampus (HC),
and PMC \citep{bzdok2015subspecialization}.
%
In line with these connectivity findings in animals and humans,
the vmPFC is often proposed to represent triggered
emotional and motivational states \citep{damasio1996somatic}.
Such real or imagined arousal states could be mapped in the vmPFC
as a bioregulatory disposition influencing cognition
and decision making.
In neuroeconomic studies of human decision making,
the vmPFC consistently reflects an individual’s subjective
value estimates
\citep{behrens2008associative}.
This may be why performance within and across participants
was related to state encoding in the vmPFC \citep{Schuck2016}.
Such a ``cognitive map'' of the action space was argued to encode
the current task state even when states are unobservable from the sensory environment.

\subsection{The hippocampus: memory, space, and experience replay}
The DMN midline has close functional links
with the HC in the medial temporal lobe \citep{vincet2006, shannon2013morning} \textemdash
a region long known to be involved in
memory operations and spatial navigation in animals and humans.

@@

The TPJ in the right-hemispheric DMN (RTPJ)
has been shown to be closely related to
multi-sensory prediction and prediction error signaling.
It is probably central for
action initiation during goal-dircted psychological tasks and for
sensorimotor behavior by integrating multi-sensory attention
\citep{corbetta2002control}.
Involvement of this DMN node was repeatedly reported in
multi-step action execution \citep{hartmann2005takes},
visuo-proprioceptive conflict \citep{Balslev2005}, and
detection of environmental changes across
visual, auditory, or tactile stimulation in an fMRI experiment
\citep{downar2000multimodal}.
Direct electrical stimulation of the human
RTPJ during neurosurgery was associated with altered perception
and stimulus awareness \citep{blanke2002neuropsychology}.
%
It was argued that the RTPJ encodes actions and ensuing outcomes,
without necessarily relating those to value estimation
\citep{liljeholm2013neural, hamilton2008action,
jakobs2009effects}.
Additionally, neural activity in the RTPJ has been proposed to reflect
stimulus-driven attentional reallocation to
self-relevant and unexpected sources of information
as a ‘‘circuit breaker’’ that recalibrates control and maintenance brain networks
\citep{bzdok2013tpj, corbettashul2008}.
% (Bzdok et al., 2013; Corbetta et al., 2008).
Indeed, patients with RTPJ damage have particular difficulties
with multi-step actions \citep{hartmann2005takes}.
In the face of large discrepancies between actual and previously predicted
environmental events the RTPJ acts as a potential switch between
externally-oriented mind sets focussed on the
sensory environment and internally-oriented mind sets focussed
on self-relevant mental scene construction.
For instance, temporally induced RTPJ damage in humans diminished the
impact of predicted intentions of other individuals
\citep{young2010disruption},
a capacity believed to be enabled by the DMN.
The RTPJ might hence be an important relay that shifts away
from the ‘‘internally directed’’ baseline processes
to, instead, deal with (unexpected) environmental stimuli and events.
% (TODO: slow endogeneous changes in state transitions by the policy and value matrices)

The TPJ in the left-hemispheric DMN (LTPJ),
in turn, has a close relationship to Wernicke's area
involved in semantic processes, such as in spoken and written language.
Neurological patients with damage in Wernicke's area
have a major impairment of language comprehension
when listening to others or reading a book.
Patient speech
preserves natural rhythm and normal syntax, yet the
voiced sentences lack meaning (i.e., aphasia).
Abstracting from speech interpretations in linguistics
and neuropsychology,
the LTPJ probably mediates access to and integration of world knowlege,
such as required during action considerations
\citep{binder2011neurobiology, seghier2013angular}.
For instance, LTPJ damage in humans also entails problems in recognizing
others' pantomimed action towards objects
without obvious relation to processing explicit language content
\citep{varney1987locus}.
% (Varney and Damasio 1987; Rothi et al. 1991; Buxbaum et al. 2005).
%
Inner speech also hinges on knowledge recall
about the physical and social world.
Indeed,
the internal production of
verbalized thought (``language of the mind'') was closely related to the LTPJ
in a pattern analysis of brain volume
\citep{geva2011neural}.
Further,
episodic memory recall and mental imagery strongly draw on
reassembling world knowledge.
Isolated building blocks of world structure probably get rebuilt
in internally constructed mental scenarios that
guide present action choice,
weigh hypothetical possibilities, and forecast the future.
%
The LTPJ may hence facilitate the automated environmental predictions
by incorporating experience-derived building blocks of world regularities
into ongoing action, planning, and problem solving.
% world -> policy matrix driven computation
% self -> value matrix driven computation

@@

\begin{wrapfigure}{r}{.4\linewidth}
  \caption{\textbf{Reinforcement learning in a nutshell.} Given the current state of the environment,
    the agent takes an action (by following some policy). This yields a reward and an observed
    state transition of the environment. The process goes on until interrupted or a goal state is reached.}
  \centering
  \includegraphics[width=1.\linewidth]{rl.pdf}
  \label{fig:rl}
\end{wrapfigure}

@@

In artificial intelligence and machine learning, a popular computational model for
multi-step decision processes in such an environment are
\textit{Markov decision processes} (MDPs) \citep{sutton1998reinforcement}.
% For the remainder of this paper, for the sake of simplicity
% we use the abbreviation ``MDP'' to refer to the special case
% of \textit{Partially Observable Markov Decision Processes}.
An MDP operationalizes a sequential decision process
in which it is assumed that environment dynamics are determined by a Markov process,
but the agent cannot directly observe the underlying state.
Instead, the agent tries to optimize a \textit{subjective} reward
signal (i.e., is likely to be different for another agent in the same state).
This is performed alongside each transition in the state space of the environment
 and maintains a probability distribution over the set of possible
states, based on a set of observations and observation probabilities.
This is a minimal set of assumptions that can be made about an environment,
and is characteristic of so-called \textit{model-free RL}. Model-free RL has had great success in real-world problems including robotics
\citep{abbeel2004,andrewng}, computer game play, \citep{mnih2015,silver2016mastering}, etc., just to name a few.
We propose to use such a model as formal basis to model DMN function.

@@

Since an action executed in the present moment might have repercussions in the far future, it turns out that the
quantity to optimize is not the instantaneous rewards $r(s, a)$, but a
cumulative reward estimate which takes into account expected reward
from action choices in the future.
A common approach to
modeling this accumulation is the time-discounted \textit{cumulative reward}% function
\begin{equation}
  \label{eq:cumr}
  G^\pi = \sum_{t=0}^{\infty}\gamma^{t}R_t = R_0 + \gamma R_1 + \gamma^2 R_2 + \ldots + \gamma^tR_t + \ldots
\end{equation}
This random variable\footnote{Random as it depends both on the environment's dynamic and the
  policy $\pi$ being played (which can be stochastic).}  measures the cumulative reward of
following a policy $\pi$.
%
Note that value buffering may be realized in the vmPFC.
This DMN node has direct connections to
to the NAc, known to be involved in reward evaluation,
and
the amygdala, known to be involved in evaluation of biological significance.

The goal of the RL agent is then to update this action policy in order
to maximize $G^\pi$ on average (cf. below). In \eqref{eq:cumr},
the definition of cumulative reward $G^\pi$,
the constant $\gamma$ $(0 \le \gamma < 1$) is the \textit{reward discount factor},
viewed to be characteristic for a certain agent.
On the one hand,
setting $\gamma = 0$ yields perfectly hedonistic behavior.
An agent with such a shortsighted horizon is exclusively
concerned with immediate rewards.
This is however not compatible with coordinated planning if long-term goal that is
potentially subserved by neural activity in the DMN.
% \footnote{In which case we can take $T \rightarrow +\infty$, provided the reward function $r$ is bounded.}.
% The reward discount factor
On the other hand,
setting $0 < \gamma < 1$ allows a learning process to arise.
A positive $\gamma$ can be seen as calibrating risk-seeking trait of the intelligent agent,
that is, the behavioral predispositions related to trading longer delays for higher rewards.
Such an agent puts relatively more emphasis on rewards expected in
a longer-term future. 
Precisely, rewards that are not expected to come within $\tau := 1/(1 - \gamma)$ time steps from the present point are disregarded.
This reduces the variance of expected rewards accumulated across
considered action cascades by limiting the depth of the search tree.
Given that there is more uncertainty in the farsighted future,
it is important to appreciate that a stochastic policy estimation
is more advantageous in many RL settings.
% For instance, a deterministic strategy in playing rock-paper-scissors can
% be easily exploited and a uniformly random choice is optimal.

@@

% \subsubsection{The Bellman equation:  characterizing optimal behavior}
% Q-learning (Watkins \& Dayan, 1992) generates a
% \textit{greedy policy}
\subsubsection{Optimal behavior and the Bellman equation}
